---
title: "MultivariateAnalysis"
author: "Yulia Iakovleva"
format: html
editor: visual
---

## Install CBASS

```{r install-cbass, message=FALSE}
if(!require(devtools)){
    install.packages("devtools")
}

devtools::install_github("greenjune-ship-it/CBASS", quiet = TRUE, force = TRUE)
```

## Load Packages

```{r library-packages, message=FALSE}
library(mice)
library(tidyr)
library(dplyr)
library(vegan)
library(CBASS)
library(ggplot2)
```

## Explore the Data

Prepare data

```{r prepare-data}
data <- cbass_dataset
data <- preprocess_dataset(data) %>% arrange(Temperature, Species, Condition)
```

Pivot the data to get PAM values as the table with Temperature as column names

```{r pivot-data}
pivoted_data <- pivot_wider(
  data,
  id_cols = c(Date, Site, Condition, Species, Genotype),
  names_from = Temperature, values_from = PAM)
# Display the result
head(pivoted_data)
```

Check Missing Values

```{r missing-values}
# Have a look at missing data
md.pattern(pivoted_data)
```

## Clean-up the Data

For now we can just discard NAs, but we can also implement imputation for some data.

```{r remove-nas}
# Remove columns with too much NA and then omit the rest
clean_data <- pivoted_data %>% select(-`33`, -`37`) %>% na.omit()
```

Make MDS

```{r mds-ordination}
# MDS ordination
mds <- metaMDS(clean_data[,6:ncol(clean_data)], distance = "euclidean")
mds <- as.data.frame(mds$points)
mds$Species <- clean_data$Species
mds$Site <- clean_data$Site
mds$Condition <- clean_data$Condition
```

Plot results

```{r plot-mds}
ggplot(mds, aes(x = MDS1, y = MDS2, color = Species, shape = Site, linetype = Condition)) +
  geom_point(size = 3) +
   stat_ellipse() +
  scale_color_brewer(palette = "Set2") # colorblind-friendly palette
```

OK, this is a nice picture that tells us nothing about the dependency of `PAM` from `Temperature`. And what's better for live.

## Try perMANOVA

```{r run-permanova}
permanova <- adonis2(
  clean_data[,6:ncol(clean_data)] ~ Site * Condition,
  data = clean_data,
  strata = clean_data$Species,
  method = "euclidean")
permanova
```

Check if we can apply perMANOVA

```{r make-PCoA-ordination}
dist <- vegdist(clean_data[,6:ncol(clean_data)], method  = "euclidean")
PCO <- betadisper(dist, clean_data$Species)
plot(PCO, main = "PCoA ordination")
```

```{r}
anova(PCO)
```
